Simple repression draft

### 2020-10-15 doc/resubmission/section_01_introduction.tex
implement most of RP comments from 20201015

Gene expression presides over much of the most important dynamism of living
organisms. The level of expression of batteries of different genes is altered as
a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals. As our ability to experimentally observe and measure the
dynamical processes that constitute the central dogma improves, there is an
opportunity to undertake a theory-experiment dialogue in order to sharpen our
understanding of such a fundamental biological process. One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems first
and foremost from our ignorance about what those architectures even are, with
more than 60\% of the genes even in an ostensibly well understood organism such
as {\it E. coli} having no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites of
a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Kuhlman2007, Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996,
Kepler2001, Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010,
Iyer-Biswas2009}. One route to such predictive understanding is to focus on the
simplest regulatory architecture and to push the theory-experiment dialogue to
increase the predictive power of our theoretical models~\cite{Garcia2011,
Phillips2019}. If we demonstrate that we can pass that test by successfully
predicting both the means and variance in gene expression at the mRNA level,
then that provides a more solid foundation upon which to launch into more
complex problems - for instance, some of the previously unknown architectures
uncovered in~\cite{Belliveau2018} and~\cite{Ireland2020}.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on the quantitative
dissection of transcription \cite{Oehler1994, Oehler1990}. One of the main
features of of the models we explore is that, by construction, all features
related to the microstates in which the repressor is bound to the promoter can
be separated from the microstates in which the RNA polymerase (RNAP) is bound.
From a modeling perspective, this means that some of the models can be written
as effective two-state models for which there is a rich
literature~\cite{Peccoud1995, Shahrezaei2008, Iyer-Biswas2009, Tkacik2009,
Sanchez2013, Jones2014, So2011, Munsky2012}. Here, we systematically compare the
predictions of several models with different levels of coarse graining written
in terms of thermodynamic and kinetic parameters. One goal in exploring such
coarse-grainings is to build towards the future models of regulatory response
that will be able to serve the powerful predictive role needed to take synthetic
biology from a brilliant exercise in enlightened empiricism to a rational design
framework as in any other branch of engineering. More precisely, we want
phenomenology in the sense of coarse-graining away atomistic detail, but still
retaining biophysical meaning. In particular a key question is: at this level of
coarse-graining, what microscopic details do we need to explicitly model, and
how do we figure that out? For example, do we need to worry about all or even
any of the steps that individual RNA polymerases go through each time they make
a transcript? Turning the question around, can we see any imprint of those
processes in the available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse (statistical
inferential) modeling are necessary to tackle such questions. We combine both
approaches in order to discard models that cannot empirically satisfy the main
features of experimental data. First we apply forward modeling to demonstrate
that none of the models are distinguishable at the level of mean gene
expression. We then extend the modeling to look at higher moments of the
distribution, eliminating models that do not empirically satisfy the observed
cell-to-cell variability. Finally we arrive at a minimal model on which we can
apply inverse modeling in order to infer the parameters that explain the data.

>>>>

Gene expression presides over much of the most important dynamism of living
organisms. The level of expression of batteries of different genes is altered as
a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals. As our ability to experimentally observe and measure the
dynamical processes that constitute the central dogma improves, there is an
opportunity to undertake a theory-experiment dialogue in order to sharpen our
understanding of such a fundamental biological process. One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems first
and foremost from our ignorance about what those architectures even are, with
more than 60\% of the genes even in an ostensibly well understood organism such
as {\it E. coli} having no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites
for a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Kuhlman2007, Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996,
Kepler2001, Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010,
Iyer-Biswas2009}. One route to such predictive understanding is to focus on the
simplest regulatory architecture and to push the theory-experiment dialogue to
increase the predictive power of our theoretical models~\cite{Garcia2011,
Phillips2019}. If we demonstrate that we can pass that test by successfully
predicting both the means and variance in gene expression at the mRNA level,
then that provides a more solid foundation upon which to launch into more
complex problems - for instance, some of the previously unknown architectures
uncovered in~\cite{Belliveau2018} and~\cite{Ireland2020}.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on the quantitative
dissection of transcription \cite{Oehler1994, Oehler1990}. One of the main
features of the models we explore is that, by construction, all features related
to the microstates in which the repressor is bound to the promoter can be
separated from the microstates in which the RNA polymerase (RNAP) is bound. From
a modeling perspective, this means that some of the models can be written as
effective two-state models for which there is a rich
literature~\cite{Peccoud1995, Shahrezaei2008, Iyer-Biswas2009, Tkacik2009,
Sanchez2013, Jones2014, So2011, Munsky2012}. Here, we systematically compare the
predictions of several models with different levels of coarse graining written
in terms of thermodynamic and kinetic parameters. One goal in exploring such
coarse-grainings is to build towards the future models of regulatory response
that will be able to serve the powerful predictive role needed to take synthetic
biology from a brilliant exercise in enlightened empiricism to a rational design
framework as in any other branch of engineering. More precisely, we want
phenomenology in the sense of coarse-graining away atomistic detail, but still
retaining biophysical meaning. In particular a key question is: at this level of
coarse-graining, what microscopic details do we need to explicitly model, and
how do we figure that out? For example, do we need to worry about all or even
any of the steps that individual RNA polymerases go through each time they make
a transcript? Turning the question around, can we see any imprint of those
processes in the available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse (statistical
inferential) modeling are necessary to tackle such questions. We combine both
approaches in order to discard models that cannot empirically satisfy the main
features of experimental data. First we apply forward modeling to demonstrate
that none of the models are distinguishable at the level of mean gene
expression. We then extend the modeling to look at higher moments of the
distribution, eliminating models that do not empirically satisfy the observed
cell-to-cell variability. Finally we arrive at a minimal model on which we can
apply inverse modeling in order to infer the parameters that explain the data.

### 2020-10-15 doc/resubmission/section_02_means.tex
implement most of RP comments from 20201015

The first class of models that are the focus of the present section on
predicting the mean level of gene expression, sometimes known as thermodynamic
models, invoke the tools of equilibrium statistical mechanics to compute the
probabilities of the promoter microstates~\cite{Ackers1982, Shea1985,
Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c, Gertz2009,
Sherman2012, Saiz2013}. As seen in Figure~\ref{fig1:means_cartoons}(B), even
within the class of thermodynamic models, we can make different commitments
about the underlying microscopic states of the promoter. Model 1 considers only
two states: a state in which a repressor (with copy number $R$) binds to an
operator and a transcriptionally active state. The free energy difference
between the repressor binding the operator, i.e. a specific binding site, and
one of the $N_{NS}$ non-specific sites is given by $\Delta\varepsilon_R$ (given
in $k_BT$ units as $\beta\equiv (k_BT)^{-1}$). Model 2 expands this model to
include an empty promoter where no transcription happens, as well as a state on
which one of the $P$ RNAPs binds to the promoter with binding energy
$\Delta\varepsilon_P$. Indeed, the list of options considered here does not at
all exhaust the suite of different microscopic states we can assign to the
promoter.

@@

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
utility in dissecting transcriptional regulation is its ratiometric nature,
which removes many secondary effects that are present when making an absolute
gene expression measurement. Also, by measuring otherwise identical cells with
and without a transcription factor present, any biological noise common to both
conditions can be made to cancel away.

Figure~\ref{fig1:means_cartoons}(B) and (C) depicts a smorgasbord of
mathematicized cartoons for simple repression using both thermodynamic and
kinetic models, respectively, that have appeared in previous literature. For
each cartoon, we calculate the fold-change in mean gene expression as predicted
by that model, deferring most algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that
for all cartoons the fold-change can be written as a Fermi function 
\begin{equation}
 \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
 \label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
where the effective free energy contains two terms: the parameters $\Delta F_R$,
an effective free energy parametrizing the repressor-DNA interaction, and
$\rho$, a term derived from the level of coarse-graining used to model all
repressor-free states. In other words, the effective free energy of the Fermi
function can be written as the additive effect of the regulation given by the
repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
that lead to a transcriptional event via $\log(\rho)$ (See
Figure~~\ref{fig1:means_cartoons}(D), left panel). This implies all models
collapse to a single master curve as shown in
Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
master curve exists and discuss why at the level of the mean expression, we are
unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
of fold-changes in expression.

@@

\subsubsection{Kinetic models}
One of the main difference between models shown in
Figure~\ref{fig1:means_cartoons}(C), casted in the language of chemical master
equations, compared with the thermodynamic models discussed in the previous 
section is the probability space over which they are built. Rather than keeping
track only of the microstate of the promoter, and assuming that gene expression
is proportional to the probability of the promoter being in a certain
microstate, chemical master equation models are built on the entire probability
state of both the promoter microstate, and the current mRNA count. Therefore, in
order to compute the fold-change, we must compute the mean mRNA count on each of
the promoter microstates, and add them all together~\cite{Sanchez2013}.

Again, we consign all details of the derivation to
Appendix~\ref{sec:non_bursty}. Here we just highlight the general findings for
all five kinetic models. As already shown in Figure~\ref{fig1:means_cartoons}(C)
and (D), all the kinetic models explored can be collapsed onto the master curve.
Given that the repressor-bound state only connects to the rest of the promoter
dynamics via its binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively,
all models can effectively be separated into two categories: a single
repressor-bound state, and all other promoter states with different levels of
coarse graining. This structure then guarantees that, at steady-state, detailed
balance between these two groups is satisfied. What this implies is that the
steady-state distribution of each of the non-repressor states has the same
functional form with or without the repressor, allowing us to write the
fold-change as a product of the ratio of the binding and unbinding rates of the
promoter, and the promoter details. This is \mrm{This is based in one of the
comments from reviewer \# 2. It is empirically true for all models we explored,
but I don't know how to prove that it is a general statement for any
unimaginable complex model.}
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
intuitive way to think about these two terms is as follows: in all kinetic
models shown in Figure~1(C) the repressor-bound state can only be reached from a
single repressor-free state. The ratio of these two states --repressor-bound and
adjacent repressor-free state-- must remain the same for all models, regardless
of the details included in other promoter states if $\Delta F_R$ represents an
effective free energy of the repressor binding the DNA operator. The presence of
other states then draws probability density from the promoter being in either of
these two states, making the ratio between the repressor-bound state and
\textit{all} repressor-free states different. The log difference in this ratio
is given by $\log(\rho)$. Since model 1 and model 5 of Figure~1(C) consist of a
single repressor-free state, $\rho$ is then necessarily 1 (See
Appendix~\ref{sec:non_bursty} for further details).

>>>>

The first class of models that are the subject of the present section focus on
predicting the mean level of gene expression. These models, sometimes known as
thermodynamic models, invoke the tools of equilibrium statistical mechanics to
compute the probabilities of the promoter microstates~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Gertz2009, Sherman2012, Saiz2013}. As seen in
Figure~\ref{fig1:means_cartoons}(B), even within the class of thermodynamic
models, we can make different commitments about the underlying microscopic
states of the promoter. Model 1 considers only two states: a state in which a
repressor (with copy number $R$) binds to an operator and a transcriptionally
active state. The free energy difference between the repressor binding the
operator, i.e. a specific binding site, and one of the $N_{NS}$ non-specific
sites is given by $\Delta\varepsilon_R$ (given in $k_BT$ units with $\beta\equiv
(k_BT)^{-1}$). Model 2 expands this model to include an empty promoter where no
transcription occurs, as well as a state in which one of the $P$ RNAPs binds to
the promoter with binding energy $\Delta\varepsilon_P$. Indeed, the list of
options considered here does not at all exhaust the suite of different
microscopic states we can assign to the promoter. The essence of thermodynamic 
models is to assign a discrete set of states and to use equilibrium statistical 
mechanics to compute the probabilities of occupancy of those states.

@@

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
utility in dissecting transcriptional regulation is its ratiometric nature,
which removes many secondary effects that are present when making an absolute
gene expression measurement. Also, by measuring otherwise identical cells with
and without a transcription factor present, any biological noise common to both
conditions can be made to cancel out.

Figure~\ref{fig1:means_cartoons}(B) and (C) depicts a smorgasbord of
mathematicized cartoons for simple repression using both thermodynamic and
kinetic models, respectively, that have appeared in previous literature. For
each cartoon, we calculate the fold-change in mean gene expression as predicted
by that model, deferring most algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the
fold-change can be written as a Fermi function of the form
\begin{equation}
 \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
 \label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
where the effective free energy contains two terms: the parameters $\Delta F_R$,
an effective free energy parametrizing the repressor-DNA interaction, and
$\rho$, a term derived from the level of coarse-graining used to model all
repressor-free states. In other words, the effective free energy of the Fermi
function can be written as the additive effect of the regulation given by the
repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
that lead to a transcriptional event via $\log(\rho)$ (See
Figure~~\ref{fig1:means_cartoons}(D), left panel). This implies all models
collapse to a single master curve as shown in
Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
master curve exists and discuss why at the level of the mean expression, we are
unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
of fold-changes in expression.

@@

\subsubsection{Kinetic models}
One of the main difference between models shown in
Figure~\ref{fig1:means_cartoons}(C), cast in the language of chemical master
equations, compared with the thermodynamic models discussed in the previous
section is the probability space over which they are built. Rather than keeping
track only of the microstate of the promoter, and assuming that gene expression
is proportional to the probability of the promoter being in a certain
microstate, chemical master equation models are built on the entire probability
state of both the promoter microstate, and the current mRNA count. Therefore, in
order to compute the fold-change, we must compute the mean mRNA count on each of
the promoter microstates, and add them all together~\cite{Sanchez2013}.

Again, we consign all details of the derivation to
Appendix~\ref{sec:non_bursty}. Here we just highlight the general findings for
all five kinetic models. As already shown in Figure~\ref{fig1:means_cartoons}(C)
and (D), all the kinetic models explored can be collapsed onto the master curve.
Given that the repressor-bound state only connects to the rest of the promoter
dynamics via its binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively,
all models can effectively be separated into two categories: a single
repressor-bound state, and all other promoter states with different levels of
coarse graining. This structure then guarantees that, at steady-state, detailed
balance between these two groups is satisfied. What this implies is that the
steady-state distribution of each of the non-repressor states has the same
functional form with or without the repressor, allowing us to write the
fold-change as a product of the ratio of the binding and unbinding rates of the
promoter, and the promoter details. This results in a fold-change of the form
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
intuitive way to think about these two terms is as follows: in all kinetic
models shown in Figure~1(C) the repressor-bound state can only be reached from a
single repressor-free state. The ratio of these two states --repressor-bound and
adjacent repressor-free state-- must remain the same for all models, regardless
of the details included in other promoter states if $\Delta F_R$ represents an
effective free energy of the repressor binding the DNA operator. The presence of
other states then draws probability density from the promoter being in either of
these two states, making the ratio between the repressor-bound state and
\textit{all} repressor-free states different. The log difference in this ratio
is given by $\log(\rho)$. Since model 1 and model 5 of Figure~1(C) consist of a
single repressor-free state, $\rho$ is then necessarily 1 (See
Appendix~\ref{sec:non_bursty} for further details).

### 2020-10-15 doc/resubmission/section_03_beyond_means.tex
implement most of RP comments from 20201015

In this section, our objective is to explore the same models considered in the
previous section, but now with reference to the the question of how well they
describe the distribution of gene expression levels, with special reference to
the variance in these distributions. To that end, we repeat the same pattern as
in the previous section by examining the models one by one. In particular
we will focus on the Fano factor, defined as the variance/mean. This metric
serves as a powerful discriminatory tool to compare our different models to the
null model that the steady-state mRNA distribution must be Poisson, resulting a
Fano factor of one.

@@

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that they represent the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
however, there are some lingering puzzles with interpreting this supercoiling
hypothesis, so we leave it as a speculation and refrain from assigning definite
physical meaning to the two states in this model.

>>>>

In this section, our objective is to explore the same models considered in the
previous section, but now with reference to the question of how well they
describe the distribution of gene expression levels, with special reference to
the variance in these distributions. To that end, we repeat the same pattern as
in the previous section by examining the models one by one. In particular we
will focus on the Fano factor, defined as the variance/mean. This metric serves
as a powerful discriminatory tool to compare our different models to the null
model that the steady-state mRNA distribution must be Poisson, resulting a Fano
factor of one.

@@

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that they represent the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
however, there are some lingering puzzles with interpreting this supercoiling
hypothesis, so we leave it as a speculative hypothesis and refrain from
assigning definite physical meaning to the two states in this model.

### 2020-10-15 doc/resubmission/section_04_bayesian_inference.tex
implement most of RP comments from 20201015

\mrm{ I would vote to remove this paragraph completely.}
We mention one further comparison, between our inferred parameters and the work
of Chong et.\ al.~\cite{Chong2014}, which is interesting and puzzling.
Experiments in~\cite{Chong2014} convincingly argue that supercoiling accumulated
from the production of mRNA transcripts is key in setting the burstiness of mRNA
production. In their model, this supercoiling occurs on the scale of
$\sim100$~kb domains of DNA. This suggests that all genes on a given domain
should burst in synchrony, and that the difference between highly and lowly
expressed genes is the \textit{size} of transcriptional bursts, not the
\textit{time between} bursts. But here, all burst sizes we infer in
Figure~\ref{fig:constit_post_full}(C) are comparable and burst rates vary
wildly. It is not immediately clear how to square this circle. Furthermore,
Figure~7E in~\cite{Chong2014} reports values of the quantity they label
$\beta/\alpha$ and we label $k^+/k^-$ in model 4 from
Figure~\ref{fig2:constit_cartoons}. In contrast to the findings
of~\cite{Razo-Mejia2020}, Chong et.\ al.\ do not find $k^+/k^-\ll1$ for most of
the genes they consider. This begs the question: is the \textit{galK}
chromosomal locus used for the reporter constructs in~\cite{Razo-Mejia2020}
and~\cite{Jones2014} merely an outlier, or is there a deeper puzzle here waiting
to be resolved? Without more apples-to-apples data we can only speculate, and we
leave it as an intriguing open question for the field.

Despite such puzzles, our goal here is not to unravel the mysterious origins of
burstiness in transcription. Our remaining task in this work is a determination
of the physical reality of thermodynamic binding energies in
Figure~\ref{fig1:means_cartoons}, as codified by the thermodynamic-kinetic
equivalence of Eq.~\ref{eq:deltaFR_eq_noneq_equiv}. For our phenomenological
needs here model 5 in Figure~\ref{fig2:constit_cartoons}(A) is more than
adequate: the posterior distributions in Figure~\ref{fig:constit_post_full}(C)
are cleanly identifiable and the predictive checks in
Figure~\ref{figS:ppc_unreg} indicate no discrepancies between the model and the
mRNA single-molecule count data of~\cite{Jones2014}. Of the models we have
considered it is unique in satisfying both these requirements. So we will
happily use it as a foundation to build upon in the next section when we add
regulation.

@@

This likelihood is rather inscrutable. We did not find any of the known
analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
we instead resorted to numerics. One insight we found
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
values, except with $k_i$ multiplied by the fold-change
$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the ratio
$k_R^+/k_R^-$ was detectable. But for intermediate repression, the distribution
was visibly broadened with Fano factor greater than $1+b$, the value for the
corresponding constitutive case. This indicates that the repressor rates had
left an imprint on the distribution, and perhaps intuitively, this intermediate
regime occurs for values of $k_R^\pm$ comparable to the burst rate $k_i$. Put
another way, if the repressor rates are much faster or much slower than $k_i$,
then there is a timescale separation and effectively only one timescale remains,
$k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three rates in the problem
are comparable does the mRNA distribution retain detectable information about
them.

>>>>

Our remaining task in this work is a determination of the physical reality of
thermodynamic binding energies in Figure~\ref{fig1:means_cartoons}, as codified
by the thermodynamic-kinetic equivalence of Eq.~\ref{eq:deltaFR_eq_noneq_equiv}.
For our phenomenological needs here model 5 in
Figure~\ref{fig2:constit_cartoons}(A) is more than adequate: the posterior
distributions in Figure~\ref{fig:constit_post_full}(C) are cleanly identifiable
and the predictive checks in Figure~\ref{figS:ppc_unreg} indicate no
discrepancies between the model and the mRNA single-molecule count data
of~\cite{Jones2014}. Of the models we have considered it is unique in satisfying
both these requirements. As a result we will use it as a foundation to build
upon in the next section when we add regulation.

@@

This likelihood is rather inscrutable. We did not find any of the known
analytical approximations for ${_2F_1}$ useful in gaining intuition, so we
instead resorted to numerics. One insight we found was that for very strong or
very weak repression, the distribution in Eq.~\ref{eq:p_m_bursty+rep} is well
approximated by a negative binomial with burst size $b$ and burst rate $k_i$
equal to their constitutive \textit{lacUV5} values, except with $k_i$ multiplied
by the fold-change $\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again
only the ratio $k_R^+/k_R^-$ was detectable. But for intermediate repression,
the distribution was visibly broadened with Fano factor greater than $1+b$, the
value for the corresponding constitutive case. This indicates that the repressor
rates had left an imprint on the distribution, and perhaps intuitively, this
intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
$k_i$. Put another way, if the repressor rates are much faster or much slower
than $k_i$, then there is a timescale separation and effectively only one
timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
rates in the problem are comparable does the mRNA distribution retain detectable
information about them.

### 2020-10-15 doc/resubmission/section_05_discussion.tex
implement most of RP comments from 20201015

With the results of the constitutive promoter in hand, we then fix the
parameters associated with this class of promoters and use them as input for
evaluating the noise in gene expression for the simple repression motif itself.
This allows us to provide a single overarching analysis of both the constitutive
and simple repression architectures using one simple model and corresponding set
of self-consistent parameters, demonstrating not only a predictive framework,
but also reconciling the thermodynamic and kinetic views of the same simple
repression constructs. More specifically, we obtained values for the
transcription factor association and dissociation rates by performing Bayesian
inference on the full mRNA distribution for data obtained from simple-repression
promoters with varying number of transcription factors per cell and affinity of
such transcription factors for the binding site. The free energy value obtained
from these kinetic rates --computed as the log ratio of the rates-- agrees with
previous inferences performed only from mean gene expression measurements, that
assumed a thermodynamic rather than a kinetic
framework~\cite{Garcia2011a,Razo-Mejia2018}. Interestingly, to constraint the
binding and unbinding rates inferred for Figure~\ref{fig4:repressed_post_full},
all of the rates need to be simultaneously fit. This is because a single
combination of operator (repressor binding site) and aTc concentration
(repressor copy number) cannot constraint both rates independently, but only
their ratio (See Appendix~\ref{sec:bayesian} for further details).

>>>>

With the results of the constitutive promoter in hand, we then fix the
parameters associated with this class of promoters and use them as input for
evaluating the noise in gene expression for the simple repression motif itself.
This allows us to provide a single overarching analysis of both the constitutive
and simple repression architectures using one simple model and corresponding set
of self-consistent parameters, demonstrating not only a predictive framework,
but also reconciling the thermodynamic and kinetic views of the same simple
repression constructs. More specifically, we obtained values for the
transcription factor association and dissociation rates by performing Bayesian
inference on the full mRNA distribution for data obtained from simple-repression
promoters with varying number of transcription factors per cell and affinity of
such transcription factors for the binding site. The free energy value obtained
from these kinetic rates --computed as the log ratio of the rates-- agrees with
previous inferences performed only from mean gene expression measurements, that
assumed a thermodynamic rather than a kinetic
framework~\cite{Garcia2011a,Razo-Mejia2018}. Interestingly, to constrain the
binding and unbinding rates inferred for Figure~\ref{fig4:repressed_post_full},
all of the rates need to be simultaneously fit. This is because a single
combination of operator (repressor binding site) and aTc concentration
(repressor copy number) cannot constraint both rates independently, but only
their ratio (See Appendix~\ref{sec:bayesian} for further details).
